---
title: "Recommended GA4 Workflow in R"
author: "Jae Jung"
date: "2026-03-08"
format: 
    html:
        theme: cosmo
        toc: true
        toc-depth: 3
        toc-location: right-body
        embed-resources: true
execute:
    freeze: auto
---

# GA4 Workflow for Minizing Query Cost

### 1. 💾 **Cache Data Locally (Most Important!)**

Since you're already familiar with DuckDB and Arrow, this is your best strategy:

```{r}
library(bigrquery)
library(dplyr)
library(arrow)

# Pull data ONCE from BigQuery
bq_project <- "your-project-id"
bq_dataset <- "analytics_XXXXXXXXX" # your GA4 dataset

# Query and download
con <- dbConnect(
  bigrquery::bigquery(),
  project = bq_project,
  dataset = bq_dataset
)

# Pull a date range once
raw_data <- tbl(con, "events_20240101") |>
  collect() # downloads to local R memory

# Save locally as Parquet (efficient, compressed)
write_parquet(raw_data, "data/ga4_events_20240101.parquet")

# Future sessions: read locally for FREE
ga4_local <- read_parquet("data/ga4_events_20240101.parquet")
```

### 2. 🦆 **Use DuckDB on Local Parquet Files**

Once cached locally, use DuckDB for fast, free querying:

```{r}
library(duckdb)
library(dplyr)

con_duck <- dbConnect(duckdb())

# Query your local parquet files — completely free!
tbl(con_duck, "read_parquet('data/ga4_events_*.parquet')") |>
  filter(event_name == "page_view") |>
  count(event_name)
```

### 3. 📅 **Query Only the Date Partitions You Need**

GA4 BigQuery tables are **date-partitioned** (e.g., `events_20240101`). Query specific dates rather than the full dataset:

```{r}
# ✅ CHEAPER: Query specific date partition
tbl(con, "events_20240101")

# ❌ MORE EXPENSIVE: Querying across all dates
# SELECT * FROM `events_*`  -- scans everything!
```

### 4. 🔍 **Select Only Columns You Need**

BigQuery charges by data scanned, so selecting fewer columns = lower cost:

```{r}
# ✅ Cheaper - only pull what you need
tbl(con, "events_20240101") |>
  select(event_name, event_timestamp, user_pseudo_id) |>
  collect()

# ❌ More expensive
tbl(con, "events_20240101") |>
  collect() # pulls all nested columns including heavy event_params
```

### 5. 🧪 **Preview Costs Before Running Queries**

Use the BigQuery console or `bq_perform_query()` with dry run to estimate costs:

```{r}
# Check estimated bytes before running
bq_dataset_query(
  bq_dataset(bq_project, bq_dataset),
  query = "SELECT * FROM events_20240101 LIMIT 10",
  dryRun = TRUE # estimates cost without charging you
)
```

## Summary

Recommended Workflow for Your Setup

```         
BigQuery (GA4 source)
      ↓  [Pull ONCE with bigrquery]
Local Parquet files (Arrow format)
      ↓  [Query freely with DuckDB]
Analysis in Quarto/R (Positron)
```

## Quick Cost Estimation

Quick Cost Estimation

| Action                                | Approx Cost                   |
|---------------------------------------|-------------------------------|
| Query 1 day of GA4 events (\~500MB)   | \~\$0.0025 (essentially free) |
| Query 1 year of GA4 events (\~180GB)  | \~\$0.90                      |
| Query 1 year, all columns, large site | Could be \$5–\$50+            |
| Reading local Parquet in DuckDB       | **\$0.00**                    |

# Analyzing GA4 BigQuery Data in R with Positron IDE

## Understanding the Challenge

GA4 data in BigQuery is notoriously complex. Each row represents a single user session/event, but fields like `event_params`, `user_properties`, and `items` are stored as **repeated RECORD types** — essentially arrays of key-value structs nested inside each row. This makes it feel intimidating, but R has excellent tools for flattening and wrangling this structure without writing raw SQL.

## 1. R Packages for BigQuery Access

### Primary Recommendation: `{bigrquery}` + `{dbplyr}`

This combination is the most idiomatic R approach and lets you write **dplyr verbs instead of SQL**.

```{r}
#| eval = false
install.packages(c("bigrquery", "dbplyr", "DBI", "dplyr"))
```

| Package     | Role                                              |
|-------------|---------------------------------------------------|
| `bigrquery` | Handles BigQuery authentication and connection    |
| `DBI`       | Provides the standard database interface layer    |
| `dbplyr`    | Translates dplyr verbs into SQL behind the scenes |

### Alternative: `{googleCloudStorageR}` + `{gargle}`

For more complex authentication scenarios (e.g., service accounts in CI/CD pipelines), `{gargle}` handles OAuth2 and service account credentials cleanly and integrates with `{bigrquery}` automatically.

```{r}
#| eval = false
install.packages(c("gargle", "googleCloudStorageR"))
```

## 2. Setting Up Your Connection

### Authentication

```{r}
#| eval = false
library(bigrquery)
library(DBI)
library(dplyr)

# Option A: Interactive OAuth (simplest for local development in Positron)
# bigrquery will open a browser window automatically on first run
bq_auth()

# Option B: Service account key file (better for reproducible/shared work)
bq_auth(path = "path/to/your/service-account-key.json")

# Option C: Use Application Default Credentials (ADC) - great for GCP environments
# Run in terminal first: gcloud auth application-default login
bq_auth(use_oob = TRUE) # Use this in Positron if browser redirect fails
```

::: callout-tip
**Positron-specific tip**: If the OAuth browser redirect fails in Positron's integrated terminal, set `use_oob = TRUE` in `bq_auth()` to get a copyable URL instead of an auto-redirect.
:::

### Establishing the Connection

```{r}
#| eval = false
# Replace with your actual project/dataset details
PROJECT_ID <- "your-gcp-project-id"
DATASET <- "analytics_XXXXXXXXX" # Your GA4 dataset ID from BigQuery

con <- dbConnect(
  bigrquery::bigquery(),
  project = PROJECT_ID,
  dataset = DATASET,
  billing = PROJECT_ID # Project that gets billed for the query
)

# Confirm the connection works by listing available tables
dbListTables(con)
# You'll see tables like: "events_20240101", "events_20240102", etc.

```

## 3. Accessing GA4 Tables with dplyr (No SQL Required)

GA4 creates one table per day (`events_YYYYMMDD`). Use `dplyr::tbl()` to create a lazy reference:

```{r}
# Point to a specific day's table
events_tbl <- tbl(con, "events_20240115")

# Or use a wildcard to span date ranges via a SQL table name trick
# This is the ONE place a small SQL string is unavoidable with bigrquery
events_range <- tbl(
  con,
  dbplyr::sql(
    "SELECT * FROM `your-project.analytics_XXXXXXXXX.events_20240101`"
  )
)
```

Better Date Range Approach Using `bq_table_download()`

```{r}
#| eval = false
library(bigrquery)

# Query multiple days using bq_project_query for more control
query_result <- bq_project_query(
  PROJECT_ID,
  query = glue::glue(
    "
    SELECT *
    FROM `{PROJECT_ID}.{DATASET}.events_*`
    WHERE _TABLE_SUFFIX BETWEEN '20240101' AND '20240131'
    LIMIT 10000
  "
  )
)

# Download results as a tibble
raw_data <- bq_table_download(query_result, bigint = "integer64")
```

::: callout-note
Even this minimal SQL is just for table selection — all the actual data transformation happens in R afterward.
:::

## 4. Packages for Handling GA4's Nested Data

Once data is pulled into R, the nested columns (`event_params`, `user_properties`, `items`) will appear as **list-columns** — columns where each cell contains a tibble or list. Here's your toolkit:

### Core Packages

```{r}
# install.packages(c("tidyr", "purrr", "tibble", "dplyr", "glue"))
```

| Package | Key Function(s) | Use Case |
|------------------------|------------------------|------------------------|
| `tidyr` | `unnest()`, `unnest_wider()`, `unnest_longer()`, `hoist()` | Flattening nested list-columns |
| `purrr` | `map()`, `map_df()`, `pluck()`, `keep()` | Iterating over list elements |
| `tibble` | `tibble()`, `enframe()` | Working with list-columns |
| `jsonlite` | `fromJSON()`, `flatten()` | If data arrives as JSON strings |

## 5. Practical Workflow: Unnesting GA4 Data

### What the Raw Data Looks Like

When you download GA4 data, `event_params` looks like this in R:

```         
# A tibble: 3 × 3
  event_name      event_params
  <chr>           <list>
1 page_view       <tibble [8 × 4]>
2 scroll          <tibble [5 × 4]>
3 click           <tibble [6 × 4]>
```

Each cell in `event_params` is itself a tibble with columns like:

-   `key` (parameter name, e.g., `"page_location"`, `"session_id"`)

-   `value.string_value`

-   `value.int_value`

-   `value.float_value`

-   `value.double_value`

### Step-by-Step Unnesting Workflow

```{r}
library(dplyr)
library(tidyr)
library(purrr)

# ── Step 1: Pull a manageable sample from BigQuery ───────────────────────────
raw_data <- bq_project_query(
  PROJECT_ID,
  query = glue::glue(
    "
    SELECT
      event_date,
      event_timestamp,
      event_name,
      event_params,
      user_pseudo_id,
      user_properties,
      device,
      geo
    FROM `{PROJECT_ID}.{DATASET}.events_*`
    WHERE _TABLE_SUFFIX = '20240115'
      AND event_name IN ('page_view', 'purchase', 'add_to_cart')
  "
  )
) |>
  bq_table_download()

# ── Step 2: Unnest event_params into long format ─────────────────────────────
params_long <- raw_data |>
  select(event_date, event_name, user_pseudo_id, event_params) |>
  unnest(event_params) # Expands list-column into rows

# Result: one row per parameter per event
# Columns: event_date, event_name, user_pseudo_id, key,
#          value.string_value, value.int_value, etc.

# ── Step 3: Consolidate the value columns into one ───────────────────────────
params_clean <- params_long |>
  mutate(
    value = coalesce(
      value.string_value,
      as.character(value.int_value),
      as.character(value.float_value),
      as.character(value.double_value)
    )
  ) |>
  select(event_date, event_name, user_pseudo_id, key, value)

# ── Step 4: Pivot to wide format (one column per parameter) ──────────────────
params_wide <- params_clean |>
  pivot_wider(
    names_from = key,
    values_from = value,
    values_fn = first # Handle duplicates by taking first value
  )

# Now you have clean columns like: page_location, session_id, ga_session_number
head(params_wide)
```

### Extracting Specific Parameters with `hoist()` (More Efficient)

If you only need a few specific parameters, `tidyr::hoist()` is faster than full unnesting because it avoids expanding everything:

```{r}
# Extract only the parameters you need without full unnest
targeted_data <- raw_data |>
  hoist(
    event_params,
    page_location = list("key" == "page_location", "value", "string_value"),
    session_id = list("key" == "ga_session_id", "value", "int_value"),
    engagement_time = list(
      "key" == "engagement_time_msec",
      "value",
      "int_value"
    )
  )
```

> **Note**: `hoist()` with named list conditions works in recent versions of `tidyr`. For older versions, use a `purrr::map()` approach shown below.

### The `purrr` Approach for Custom Extraction

This is more verbose but gives you maximum control:

```{r}
# Helper function to extract a specific param value from the nested list
extract_param <- function(params_list, param_key) {
  purrr::map_chr(params_list, function(params) {
    # params is a tibble with key/value columns
    match_row <- params[params$key == param_key, ]

    if (nrow(match_row) == 0) {
      return(NA_character_)
    }

    # Return whichever value type is populated
    coalesce(
      match_row$value.string_value,
      as.character(match_row$value.int_value),
      as.character(match_row$value.float_value)
    )[1]
  })
}

# Apply to your dataframe
enriched_data <- raw_data |>
  mutate(
    page_location = extract_param(event_params, "page_location"),
    session_id = extract_param(event_params, "ga_session_id"),
    page_referrer = extract_param(event_params, "page_referrer")
  ) |>
  select(-event_params) # Drop the now-redundant list column
```

## 6. Handling Other Nested Structures

### Device and Geo (Struct columns — easier!)

These are nested structs, not repeated records, so they unnest cleanly:

```{r}
device_data <- raw_data |>
  unnest_wider(device, names_sep = "_") |> # Creates device_category, device_browser, etc.
  unnest_wider(geo, names_sep = "_") # Creates geo_country, geo_city, etc.
```

### User Properties

Same pattern as `event_params`:

```{r}
user_props <- raw_data |>
  select(user_pseudo_id, user_properties) |>
  unnest(user_properties) |>
  mutate(value = coalesce(value.string_value, as.character(value.int_value))) |>
  select(user_pseudo_id, key, value) |>
  pivot_wider(names_from = key, values_from = value, values_fn = first)
```

### E-commerce Items Array

```{r}
purchase_items <- raw_data |>
  filter(event_name == "purchase") |>
  select(user_pseudo_id, event_timestamp, items) |>
  unnest(items) |> # One row per item per purchase
  unnest_wider(item_params) # Expand any nested item params
```

## 7. Complete End-to-End Example

Here's a realistic workflow for a common GA4 analysis — **page view analysis by device**:

```{r}
library(bigrquery)
library(DBI)
library(dplyr)
library(tidyr)
library(purrr)
library(glue)
library(lubridate)

# ── Setup ─────────────────────────────────────────────────────────────────────
PROJECT_ID <- "your-project-id"
DATASET <- "analytics_XXXXXXXXX"
bq_auth()

# ── Pull page view events ─────────────────────────────────────────────────────
raw <- bq_project_query(
  PROJECT_ID,
  glue(
    "
    SELECT
      event_date,
      event_name,
      user_pseudo_id,
      event_params,
      device.category    AS device_category,
      device.browser     AS browser,
      geo.country        AS country
    FROM `{PROJECT_ID}.{DATASET}.events_*`
    WHERE _TABLE_SUFFIX BETWEEN '20240101' AND '20240131'
      AND event_name = 'page_view'
  "
  )
) |>
  bq_table_download()

# ── Extract page location from nested params ───────────────────────────────────
page_views <- raw |>
  mutate(
    # Use purrr to safely extract the page_location parameter
    page_location = map_chr(event_params, function(p) {
      val <- p$value.string_value[p$key == "page_location"]
      if (length(val) == 0) NA_character_ else val[1]
    }),
    event_date = ymd(event_date)
  ) |>
  select(-event_params) |> # Clean up
  filter(!is.na(page_location)) |>
  mutate(
    # Parse path from full URL
    page_path = gsub("https?://[^/]+", "", page_location),
    page_path = gsub("\\?.*", "", page_path) # Remove query strings
  )

# ── Analyze ────────────────────────────────────────────────────────────────────
# Top pages by device
top_pages_by_device <- page_views |>
  count(device_category, page_path, sort = TRUE) |>
  group_by(device_category) |>
  slice_max(n, n = 10) |>
  ungroup()

# Daily trend
daily_trend <- page_views |>
  count(event_date, device_category) |>
  arrange(event_date)

print(top_pages_by_device)
print(daily_trend)
```

## 8. Best Practices & Pro Tips

### Performance

-   **Always filter in the query first**: Pull only the event types, date ranges, and columns you need. GA4 tables can be enormous.

-   **Use `LIMIT` during development**: Add `LIMIT 1000` while building your pipeline, remove it for final runs.

-   **Cache downloaded data locally**: Save intermediate results with `readr::write_rds()` to avoid re-querying BigQuery during development.

```{r}
# Cache pattern
cache_file <- "data/raw_ga4_jan2024.rds"

if (!file.exists(cache_file)) {
  raw_data <- bq_project_query(...) |> bq_table_download()
  readr::write_rds(raw_data, cache_file, compress = "gz")
} else {
  raw_data <- readr::read_rds(cache_file)
}
```

### Working in Positron

-   Use Positron's **Variables pane** to inspect list-column structure interactively — click on a cell to see the nested tibble without printing to console.

-   The **integrated data viewer** handles tibbles with list-columns better than RStudio in many cases.

-   Create a dedicated **R Project** with a `renv.lock` file to pin package versions, especially `bigrquery` (it updates frequently).

```{r}
# Set up reproducible environment
install.packages("renv")
renv::init()
renv::snapshot() # After installing your packages
```

### GA4-Specific Tips

| Tip | Detail |
|----------------------|--------------------------------------------------|
| **`user_pseudo_id` ≠ user** | It resets if cookies are cleared; use with caution for user counts |
| **Deduplicate events** | GA4 can have duplicate events; use `distinct()` on `event_timestamp + user_pseudo_id + event_name` |
| **Timezone awareness** | `event_timestamp` is in microseconds UTC; convert with `as.POSIXct(event_timestamp / 1e6, origin = "1970-01-01", tz = "America/New_York")` |
| **Intraday tables** | Same-day data is in `events_intraday_YYYYMMDD` — query separately if you need today's data |

### Debugging Nested Columns

```{r}
# Peek inside a list-column without unnesting everything
raw_data$event_params[[1]] # View the tibble for the first row
raw_data$event_params[[1]] |> View() # Open in Positron's data viewer

# Check what keys exist across all events
raw_data |>
  unnest(event_params) |>
  count(key, sort = TRUE) # Inventory of all parameter names
```

### Quick Reference Card

```         
BigQuery Connection:    bigrquery + DBI + dbplyr
Authentication:         bq_auth() / service account JSON
Download data:          bq_table_download()
Unnest repeated RECORD: tidyr::unnest() → pivot_wider()
Unnest struct columns:  tidyr::unnest_wider(col, names_sep="_")
Selective extraction:   purrr::map_chr() with custom helper
Consolidate values:     dplyr::coalesce(str_val, as.character(int_val), ...)
Cache results:          readr::write_rds() / read_rds()
```

::: callout-note
This workflow keeps you almost entirely in the tidyverse, with BigQuery handling only the initial data retrieval — exactly the right division of labor for an R-focused analysis workflow.
:::