---
title: "GA4 BigQuery Analysis with Arrow & DuckDB"
subtitle: "Demonstration of Efficient large-scale analytics in R using Positron"
description: "This page shows how to work with GA4 data using Arrow & DuckDB"
categories: [BigQuery, Arrow, Parquet, DuckDB, SQL]
order: 2
author: "Jae Jung"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    theme: cosmo
    highlight-style: github
    embed-resources: true
execute:
  echo: fenced
  warning: false
  message: false
  cache: true  # Cache chunks to avoid re-running expensive queries
---

# Setup

## Install & Load Packages

```{r}
#| label: install-packages
#| eval: false

# Run once to install required packages
install.packages(c(
  "bigrquery", # BigQuery connector
  "DBI", # Database Interface
  "dplyr", # Data manipulation
  "arrow", # Apache Arrow (columnar, efficient)
  "duckdb", # In-process OLAP SQL engine
  "ggplot2", # Plotting
  "lubridate", # Date handling
  "scales", # Axis formatting
  "glue" # String interpolation
))
```

```{r}
#| label: load-packages

library(bigrquery)
library(DBI)
library(dplyr)
library(arrow)
library(duckdb)
library(tidyverse)
```

------------------------------------------------------------------------

# 1. Connect to BigQuery

GA4 exports data to BigQuery in sharded daily tables named `events_YYYYMMDD`. The public dataset `bigquery-public-data.ga4_obfuscated_sample_ecommerce` is a great sandbox — no credentials required beyond a GCP project.

```{r}
#| label: bigquery-connection

# ── Configuration ────────────────────────────────────────────────────────────
GCP_PROJECT <- "ga4-bigquery-cppmsdm" # <-- change this
BQ_PROJECT <- "bigquery-public-data"
BQ_DATASET <- "ga4_obfuscated_sample_ecommerce"

# Authenticate (opens browser on first run; token cached in ~/.cache/gargle)
# bq_auth()                          # uncomment if not yet authed

# bq_user()

# Open a DBI connection to BigQuery
bq_con <- dbConnect(
  bigrquery::bigquery(),
  project = BQ_PROJECT,
  dataset = BQ_DATASET,
  billing = GCP_PROJECT
)

# Confirm tables are visible
dbListTables(bq_con)
```

------------------------------------------------------------------------

# 2. Pull Data from BigQuery → Arrow Table

Rather than pulling everything into memory, we:

1.  Write a focused SQL query in BigQuery (push-down filtering)
2.  Stream results directly into an **Arrow Table** (columnar, zero-copy)
3.  Hand it off to **DuckDB** for further in-process analytics

## 2a. Date range helper

```{r}
#| label: date-range

# Define a rolling 90-day window
#date_end   <- Sys.Date() - 1                         # yesterday
#date_start <- date_end - 89                          # 90 days back

date_end <- as.Date("2021-01-31")
date_start <- as.Date("2021-01-01")

# Format as YYYYMMDD for BQ table suffix pattern
bq_start <- format(date_start, "%Y%m%d")
bq_end <- format(date_end, "%Y%m%d")

```

## 2b. BigQuery SQL — flatten GA4 nested structs

> GA4 data is heavily nested (events → event_params, items, etc.). We unnest the key fields we need in SQL before pulling.

```{r}
#| label: bq-sql
library(glue)

sql <- glue(
  "
  SELECT
    -- Session / user dims
    PARSE_DATE('%Y%m%d', event_date) AS event_date,
    user_pseudo_id,
    (SELECT value.int_value
       FROM UNNEST(event_params)
      WHERE key = 'ga_session_id') AS session_id,

    -- Traffic source
    traffic_source.source AS source,
    traffic_source.medium AS medium,
    traffic_source.name AS campaign,

    -- Device
    device.category AS device_category,
    device.operating_system AS os,

    -- Geo
    geo.country AS country,

    -- Event
    event_name,
    event_timestamp,

    -- Ecommerce (NULL for non-purchase events)
    ecommerce.purchase_revenue AS revenue,
    ecommerce.transaction_id AS transaction_id

  FROM `{BQ_PROJECT}.{BQ_DATASET}.events_*`
  WHERE _TABLE_SUFFIX BETWEEN '{bq_start}' AND '{bq_end}'
    AND event_name IN (
      'session_start', 'page_view',
      'view_item', 'add_to_cart', 'purchase'
    )
"
)

cat(sql)
```

## 2c. Execute & capture as Arrow Table

```{r}
#| label: fetch-arrow
#| eval: false
#| cache: true

# Execute the query; stream results into an Arrow Table
# arrow::read_ipc_stream() handles memory efficiently
result <- DBI::dbGetQuery(bq_con, sql)
```

::: callout-warning
> Unfortunately, this code above led to the following error when executed.

```         
Billed: 362.81 MB
Downloading first chunk of data.
Downloading the remaining 629,250 rows in 63 chunks of (up to) 10,000 rows.
                                                              
Warning messages:
1: In as.integer.integer64(x) : NAs produced by integer overflow
2: In as.integer.integer64(x) : NAs produced by integer overflow
```

> Fixing the problem with requring with the following way.
:::

```{r}
# use bigrquery functions to control bigint conversion
tbl <- bigrquery::bq_project_query(GCP_PROJECT, sql) # billed 362.81 MB
result <- bigrquery::bq_table_download(
  tbl,
  bigint = "numeric",
  page_size = 10000
)
result # A tibble: 639,250 × 13
```

> Continue...

```{r}
# Convert to Arrow Table (columnar, faster for downstream DuckDB queries)
ga4_arrow <- arrow::as_arrow_table(result) # scannable Arrow object

# Disconnect from BigQuery — we don't need it anymore
dbDisconnect(bq_con)

# Quick look
print(ga4_arrow)
```

> **Tip — very large datasets:** If the dataset exceeds available RAM, write it to Parquet first and query the files directly with `DuckDB`:
>
> ``` r
> arrow::write_parquet(ga4_arrow, "ga4_data.parquet")
> # Then use duckdb::duckdb_register_arrow() or read_parquet() in SQL
> ```

------------------------------------------------------------------------

# 3. Register Arrow Table in DuckDB

> DuckDB can query Arrow Tables **without copying data** via zero-copy integration — ideal for large datasets.

```{r}
#| label: duckdb-setup

# In-memory DuckDB instance (no file needed)
duck <- dbConnect(duckdb::duckdb())

# Register the Arrow table as a virtual DuckDB view
duckdb::duckdb_register_arrow(duck, "ga4", ga4_arrow)

# Confirm
dbListTables(duck)
```

-   Check to confirm and list what's available

```{r}
#| eval: true
# Confirm Arrow object exists and type
exists("ga4_arrow")
class(ga4_arrow)
print(ga4_arrow)

# Confirm registration of arrow as a virtual DuckDB by querying the table directly
DBI::dbGetQuery(duck, "SELECT COUNT(*) AS n FROM ga4")

# List SQL-visible tables (information_schema is more reliable than SHOW TABLES)
DBI::dbGetQuery(
  duck,
  "
  SELECT table_schema, table_name
  FROM information_schema.tables
  ORDER BY table_schema, table_name
"
)
```

**Output**:

```         
> exists("ga4_arrow")
[1] TRUE
> class(ga4_arrow)
[1] "Table"        "ArrowTabular" "ArrowObject"  "R6"          
> print(ga4_arrow)
Table
639250 rows x 13 columns
$event_date <date32[day]>
$user_pseudo_id <string>
$session_id <double>
$source <string>
$medium <string>
$campaign <string>
$device_category <string>
$os <string>
$country <string>
$event_name <string>
$event_timestamp <double>
$revenue <double>
$transaction_id <string>
> DBI::dbGetQuery(duck, "SELECT COUNT(*) AS n FROM ga4")
       n
1 639250
> DBI::dbGetQuery(duck, "
+   SELECT table_schema, table_name
+   FROM information_schema.tables
+   ORDER BY table_schema, table_name
+ ")
                     
  table_schema  table_name
1         main ga4_parquet
```

------------------------------------------------------------------------

# 4. Analysis with DuckDB SQL

## 4a. Daily sessions & revenue

```{r}
#| label: daily-sessions
#| eval: true

daily <- dbGetQuery(
  duck,
  "
  SELECT
    event_date,
    COUNT(DISTINCT CASE WHEN event_name = 'session_start'
                        THEN user_pseudo_id || CAST(session_id AS VARCHAR)
                   END)                           AS sessions,
    COUNT(DISTINCT CASE WHEN event_name = 'purchase'
                        THEN transaction_id END)  AS transactions,
    ROUND(SUM(COALESCE(revenue, 0)), 2)           AS revenue
  FROM ga4
  GROUP BY event_date
  ORDER BY event_date
"
) |>
  mutate(event_date = as.Date(event_date))

head(daily)
```

**Output**:

```         
> head(daily)
  event_date sessions transactions revenue
1 2021-01-01     2289           13     975
2 2021-01-02     2705           14    1026
3 2021-01-03     2753           14     552
4 2021-01-04     3317           16     834
5 2021-01-05     4410           28    1143
6 2021-01-06     5951           31    2000
```

```{r}
#| label: plot-daily
#| fig-width: 10
#| fig-height: 4
#| eval: true

library(scales)

ggplot(daily, aes(x = event_date)) +
  geom_col(aes(y = sessions), fill = "#4285F4", alpha = 0.8) +
  geom_line(
    aes(y = revenue * max(sessions) / max(revenue)),
    color = "#EA4335",
    linewidth = 1
  ) +
  scale_y_continuous(
    name = "Sessions",
    labels = comma,
    sec.axis = sec_axis(
      ~ . * max(daily$revenue) / max(daily$sessions),
      name = "Revenue (USD)",
      labels = dollar
    )
  ) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week") +
  labs(title = "Daily Sessions & Revenue", x = NULL) +
  theme_minimal(base_size = 13) +
  theme(axis.title.y.right = element_text(color = "#EA4335"))
```

## 4b. Channel performance

```{r}
#| label: channel-performance

channels <- dbGetQuery(
  duck,
  "
  WITH sessions AS (
    SELECT
      source,
      medium,
      user_pseudo_id || CAST(session_id AS VARCHAR) AS session_key
    FROM ga4
    WHERE event_name = 'session_start'
  ),
  purchases AS (
    SELECT
      source,
      medium,
      COUNT(DISTINCT transaction_id) AS transactions,
      SUM(revenue)                   AS revenue
    FROM ga4
    WHERE event_name = 'purchase'
    GROUP BY source, medium
  )
  SELECT
    s.source,
    s.medium,
    COUNT(DISTINCT s.session_key)         AS sessions,
    COALESCE(p.transactions, 0)           AS transactions,
    ROUND(COALESCE(p.revenue, 0), 2)      AS revenue,
    ROUND(100.0 * COALESCE(p.transactions, 0)
          / NULLIF(COUNT(DISTINCT s.session_key), 0), 2) AS cvr_pct
  FROM sessions s
  LEFT JOIN purchases p USING (source, medium)
  GROUP BY s.source, s.medium, p.transactions, p.revenue
  ORDER BY sessions DESC
  LIMIT 20
"
)

channels
```

## 4c. Funnel analysis

```{r}
#| label: funnel

all_counts <- dbGetQuery(
  duck,
  "
  SELECT
    event_name,
    COUNT(DISTINCT user_pseudo_id) AS users
  FROM ga4
  GROUP BY event_name
"
)

funnel <- all_counts |>
  filter(
    event_name %in%
      c(
        "session_start",
        "page_view",
        "view_item",
        "add_to_cart",
        "purchase"
      )
  ) |>
  mutate(
    event_name = factor(
      event_name,
      levels = c(
        "session_start",
        "page_view",
        "view_item",
        "add_to_cart",
        "purchase"
      )
    ),
    step_pct = round(100 * users / max(users), 1)
  ) |>
  arrange(event_name)

funnel
```

```{r}
#| label: plot-funnel
#| fig-width: 8
#| fig-height: 4

ggplot(funnel, aes(x = event_name, y = users, fill = event_name)) +
  geom_col(show.legend = FALSE) +
  geom_text(
    aes(label = glue("{comma(users)}\n({step_pct}%)")),
    vjust = -0.3,
    size = 3.5
  ) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  scale_fill_brewer(palette = "Blues", direction = -1) +
  labs(title = "GA4 Event Funnel", x = NULL, y = "Unique Users") +
  theme_minimal(base_size = 13)
```

## 4d. Device breakdown

```{r}
#| label: device-breakdown

devices <- dbGetQuery(
  duck,
  "
  SELECT
    device_category,
    COUNT(DISTINCT user_pseudo_id || CAST(session_id AS VARCHAR)) AS sessions,
    ROUND(SUM(COALESCE(revenue,0)), 2)  AS revenue
  FROM ga4
  GROUP BY device_category
  ORDER BY sessions DESC
"
)

devices
```

------------------------------------------------------------------------

# 5. Using dplyr on Arrow (alternative to raw SQL)

You can also use familiar `dplyr` verbs directly on the Arrow table. Arrow translates them to efficient columnar operations.

```{r}
#| label: dplyr-on-arrow

# dplyr on Arrow — lazy evaluation, no full materialisation until collect()
top_countries <- ga4_arrow |>
  as_tibble() |> # convert Arrow Table to a tibble
  filter(event_name == "purchase") |>
  group_by(country) |>
  summarise(
    transactions = n_distinct(transaction_id),
    revenue = sum(revenue, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(revenue)) |>
  slice_head(n = 10)

top_countries
```

------------------------------------------------------------------------

# 6. Save Results to Parquet (for later reuse)

```{r}
#| label: save-parquet
#| eval: false

# Write Arrow Table to Parquet — compressed, fast to re-read
arrow::write_parquet(
  ga4_arrow,
  "data/ga4_export.parquet",
  compression = "snappy"
)

# Later: read back without touching BigQuery at all
ga4_from_parquet <- arrow::read_parquet("data/ga4_export.parquet")

# Or query it directly from DuckDB
dbExecute(
  duck,
  "
  CREATE VIEW ga4_parquet AS
  SELECT * FROM read_parquet('data/ga4_export.parquet')
"
)
```

------------------------------------------------------------------------

# 7. Cleanup

```{r}
#| label: cleanup

# Always disconnect DuckDB when done
dbDisconnect(duck, shutdown = TRUE)
```

------------------------------------------------------------------------

# Workflow Summary

| Step | Tool | Why |
|------------------------|------------------------|------------------------|
| Auth & query BigQuery | `bigrquery` + `DBI` | Push filtering to BQ, pay only for bytes scanned |
| Store result | `arrow` Arrow Table | Columnar, zero-copy, RAM-efficient |
| Heavy analytics | `duckdb` | Fast in-process OLAP SQL, no server needed |
| Tidy transforms | `dplyr` on Arrow | Familiar syntax, lazy evaluation |
| Persist | `arrow::write_parquet()` | Avoid re-querying BQ on every run |
| Visualise | `ggplot2` | Standard R plotting |

> **Positron tip:** Use the Variables pane to inspect Arrow Tables and DuckDB query results inline. The integrated terminal lets you render this document with `quarto render ga4_bigquery_analysis.qmd`.