---
title: "Rectangling Hierarchical Data"
subtitle: "Summary of the Fundamentals of unnesting list-columns explained in Ch23, RDDS"
description: "This page summarizes the fundamental concepts about hierarchal data and how to flatten it"
categories: [tidyr, JSON, tidyverse]
order: 1
author: "Jae Jung"
date: today
format:
  html:
    theme:
      light: flatly
      dark: darkly
    toc: true
    toc-depth: 3
    toc-location: left-body
    toc-title: "Contents"
    toc-expand: true
    code-fold: show
    code-copy: true
    code-tools: true
    highlight-style: github
    fig-width: 9
    fig-height: 5
    fig-align: center
    df-print: paged
    number-sections: true
    smooth-scroll: true
    #embed-resources: true
    callout-appearance: minimal
execute:
  warning: false
  message: false
  cache: true
  freeze: auto
---

# What is data rectangling?

-   Taking data that is fundamentally hierarchical, or tree-like, and converting it into a rectangular data frame made up of rows and columns.

-   This is important because hierarchical data is surprisingly common, especially when working with data that comes from the web.

-   lists make hierarchical data possible.

    -   `tidyr::unnest_longer()` and `tidyr::unnest_wider()`.

## Packages to use

```{r}
#install.packages("repurrrsive")
library(tidyverse)
library(repurrrsive)
library(jsonlite)
```

# Lists

```{r}
x1 <- list(1:4, "a", TRUE)
x1
```

> Naming the components, or children, of a list

```{r}
x2 <- list(a = 1:2, b = 1:3, c = 1:4)
x2
```

> To print use `str()`, which generates a compact display of the structure, de-emphasizing the contents:

```{r}
str(x1)

str(x2)
```

## Hierarchy

> -   Lists can contain any type of object, including other lists.
> -   This makes them suitable for representing hierarchical (tree-like) structures:

```{r}
x3 <- list(list(1, 2), list(3, 4))
str(x3)
```

> -   This is notably different to c(), which generates a flat vector:

```{r}
c(c(1, 2), c(3, 4))
```

```{r}
x5 <- list(1, list(2, list(3, list(4, list(5)))))
str(x5)

view(x5)
```

## List-Columns

```{r}
df <- tibble(
  x = 1:2,
  y = c("a", "b"),
  z = list(list(1, 2), list(3, 4, 5))
)
df
```

```{r}
df |>
  filter(x == 1)
```

> -   Computing with list-columns is harder.

> -   tibble() treats lists like vectors and the print method has been designed with lists in mind.

# Unnesting

> -   List-columns tend to come in two basic forms: named and unnamed.

> -   When the children are named, they tend to have the same names in every row.

## Named list-columns

> -   Named list-columns naturally unnest into **columns**: each named element becomes a new named column.

```{r}
df1 <- tribble(
  ~x , ~y                   ,
   1 , list(a = 11, b = 12) ,
   2 , list(a = 21, b = 22) ,
   3 , list(a = 31, b = 32) ,
)
df1
```

## Unnamed list-columns

> -   When the children are unnamed, the number of elements tends to vary from row-to-row.

> -   Unnamed list-columns naturally unnest into **rows**: you’ll get one row for each child.

```{r}
df2 <- tribble(
  ~x , ~y               ,
   1 , list(11, 12, 13) ,
   2 , list(21)         ,
   3 , list(31, 32)     ,
)
```

## `unnest_wider()`

> Used when each row has the same number of elements with the same names

```{r}
df1 |>
  unnest_wider(y)
```

> By default, the names of the new columns come exclusively from the names of the list elements, but you can use the `names_sep` argument to request that they combine the column name and the element name. This is useful for disambiguating repeated names.

```{r}
df1 |>
  unnest_wider(y, names_sep = "_")
```

## `unnest_longer()`

> Used when each row contains an unnamed list

```{r}
df2 |>
  unnest_longer(y)
```

> Note how x is duplicated for each element inside of y: we get one row of output for each element inside the list-column.

> But what happens if one of the elements is empty, as in the following example?

```{r}
df6 <- tribble(
  ~x  , ~y         ,
  "a" , list(1, 2) ,
  "b" , list(3)    ,
  "c" , list()
)

df6 |>
  unnest_longer(y)
```

> We get zero rows in the output, so the row effectively disappears.

> If you want to preserve that row, adding `NA` in y, set `keep_empty` = `TRUE`.

```{r}
df6 <- tribble(
  ~x  , ~y         ,
  "a" , list(1, 2) ,
  "b" , list(3)    ,
  "c" , list(NA)
)

df6 |>
  unnest_longer(y, keep_empty = TRUE)
```

## Inconsitent Types

> What happens if you unnest a list-column that contains different types of vector?

```{r}
df4 <- tribble(
  ~x  , ~y                 ,
  "a" , list(1)            ,
  "b" , list("a", TRUE, 5)
)
```

```{r}
df4 |>
  unnest_longer(y)
```

> The output contains a list-column, but every element of the list-column contains a single element.

> Dealing with inconsistent types is challenging and the details depend on the precise nature of the problem and your goals, but you’ll most likely need tools from Iteration.

## Other functions from tidyr

-   [`unnest_auto()`](https://tidyr.tidyverse.org/reference/unnest_auto.html) automatically picks between [`unnest_longer()`](https://tidyr.tidyverse.org/reference/unnest_longer.html) and [`unnest_wider()`](https://tidyr.tidyverse.org/reference/unnest_wider.html) based on the structure of the list-column. It’s great for rapid exploration, but ultimately it’s a bad idea because it doesn’t force you to understand how your data is structured, and makes your code harder to understand.

-   [`unnest()`](https://tidyr.tidyverse.org/reference/unnest.html) expands both rows and columns. It’s useful when you have a list-column that contains a **2d structure** like a data frame, which you don’t see in this book, but you might encounter if you use the [tidymodels](https://www.tmwr.org/base-r.html#combining-base-r-models-and-the-tidyverse) ecosystem.

## Exercises

1.  What happens when you use [`unnest_wider()`](https://tidyr.tidyverse.org/reference/unnest_wider.html) with unnamed list-columns like `df2`? What argument is now necessary? What happens to missing values?

2.  What happens when you use [`unnest_longer()`](https://tidyr.tidyverse.org/reference/unnest_longer.html) with named list-columns like `df1`? What additional information do you get in the output? How can you suppress that extra detail?

3.  From time-to-time you encounter data frames with multiple list-columns with aligned values. For example, in the following data frame, the values of `y` and `z` are aligned (i.e. `y` and `z` will always have the same length within a row, and the first value of `y` corresponds to the first value of `z`). What happens if you apply two [`unnest_longer()`](https://tidyr.tidyverse.org/reference/unnest_longer.html) calls to this data frame? How can you preserve the relationship between `x` and `y`? (Hint: carefully read the docs).

```{r}
df4 <- tribble(
  ~x  , ~y                              , ~z                              ,
  "a" , list("y-a-1", "y-a-2")          , list("z-a-1", "z-a-2")          ,
  "b" , list("y-b-1", "y-b-2", "y-b-3") , list("z-b-1", "z-b-2", "z-b-3")
)
```

# Case Studies

> Real data typically contains multiple levels of nesting that require multiple calls to unnest_longer() and/or unnest_wider().

## Very wide data

> **gh_repos**. This is a list that contains data about a collection of GitHub repositories retrieved using the GitHub API.

```{r}
library(repurrrsive)
#view(gh_repos)

repos <- tibble(json = gh_repos)
repos
```

```{r}
repos |>
  unnest_longer(json)
```

```{r}
repos |>
  unnest_longer(json) |>
  unnest_wider(json)

```

```{r}
repos |>
  unnest_longer(json) |>
  unnest_wider(json) |>
  names() |>
  head(10)
```

```{r}
repos |>
  unnest_longer(json) |>
  unnest_wider(json) |>
  select(id, full_name, owner, description)
```

```{r}
#| eval: false
repos |>
  unnest_longer(json) |>
  unnest_wider(json) |>
  select(id, full_name, owner, description) |>
  unnest_wider(owner)
```

**Output**:

```         
Error in `unnest_wider()` at w09-hierarchical-data/ch23/draft-hierarchical-data-0.qmd:301:1:
! Can't duplicate names between the affected columns and the original data.
✖ These names are duplicated:
  ℹ `id`, from `owner`.
ℹ Use `names_sep` to disambiguate using the column name.
```

```{r}
repos |>
  unnest_longer(json) |>
  unnest_wider(json) |>
  select(id, full_name, owner, description) |>
  unnest_wider(owner, names_sep = "_")
```

## Relational data

> Nested data is sometimes used to represent data that we’d usually spread across multiple data frames.

> **Example**: `got_chars` which contains data about characters that appear in the Game of Thrones books and TV series.

```{r}
chars <- tibble(json = got_chars)
chars

# Widening
chars |>
  unnest_wider(json)

# selecting a few
chars |>
  unnest_wider(json) |>
  select(id, name, gender, culture, born, died, alive)

# search for list-columns
chars |>
  unnest_wider(json) |>
  select(id, where(is.list))

# explore titles column
chars |>
  unnest_wider(json) |>
  select(id, titles) |>
  unnest_longer(titles)

# cleanning
titles <- chars |>
  unnest_wider(json) |>
  select(id, titles) |>
  unnest_longer(titles) |>
  filter(titles != "") |>
  rename(title = titles)

titles
```

> You could imagine creating a table like this for each of the list-columns, then using joins to combine them with the character data as you need it.

## Deeply nested

> When the data is very deeply nested and requires repeated rounds of `unnest_wider()` and `unnest_longer()` to unravel: `gmaps_cities`.

```{r}
gmaps_cities
```

```{r}
gmaps_cities |>
  unnest_wider(json)
```

> -   Drop the status column since they’re all OK

> -   Unnamed list

```{r}
gmaps_cities |>
  unnest_wider(json) |>
  select(-status) |>
  unnest_longer(results)
```

> Now results is a named list, so we’ll use `unnest_wider()`:

```{r}
locations <- gmaps_cities |>
  unnest_wider(json) |>
  select(-status) |>
  unnest_longer(results) |>
  unnest_wider(results)

locations
```

> There are a few different places we could go from here. We might want to determine the exact location of the match, which is stored in the `geometry` list-column:

```{r}
locations |>
  select(city, formatted_address, geometry) |>
  unnest_wider(geometry)
```

> We can unnest location to see the latitude (lat) and longitude (lng):

```{r}
locations |>
  select(city, formatted_address, geometry) |>
  unnest_wider(geometry) |>
  unnest_wider(location)
```

> Extracting the bounds requires a few more steps:

```{r}
locations |>
  select(city, formatted_address, geometry) |>
  unnest_wider(geometry) |>
  # focus on the variables of interest
  select(!location:viewport) |>
  unnest_wider(bounds)
```

> We then rename southwest and northeast (the corners of the rectangle) so we can use names_sep to create short but evocative names:

```{r}
locations |>
  select(city, formatted_address, geometry) |>
  unnest_wider(geometry) |>
  select(!location:viewport) |>
  unnest_wider(bounds) |>
  rename(ne = northeast, sw = southwest) |>
  unnest_wider(c(ne, sw), names_sep = "_")
```

> Once you’ve discovered the path to get to the components you’re interested in, you can extract them directly using another tidyr function, hoist():

```{r}
locations |>
  select(city, formatted_address, geometry) |>
  hoist(
    geometry,
    ne_lat = c("bounds", "northeast", "lat"),
    sw_lat = c("bounds", "southwest", "lat"),
    ne_lng = c("bounds", "northeast", "lng"),
    sw_lng = c("bounds", "southwest", "lng"),
  )
```

## What Is `hoist()`?

`hoist()` is a function from `{tidyr}` that lets you **pull specific elements out of a list-column by name or position**, without unnesting the entire structure. Think of it as a surgical tool compared to the blunt instruments of `unnest_wider()` and `unnest_longer()`

## Exercises

1.  Roughly estimate when `gh_repos` was created. Why can you only roughly estimate the date?

2.  The `owner` column of `gh_repos` contains a lot of duplicated information because each owner can have many repos. Can you construct an `owners` data frame that contains one row for each owner? (Hint: does [`distinct()`](https://dplyr.tidyverse.org/reference/distinct.html) work with `list-cols`?)

3.  Follow the steps used for `titles` to create similar tables for the aliases, allegiances, books, and TV series for the Game of Thrones characters.

4.  Explain the following code line-by-line. Why is it interesting? Why does it work for `got_chars` but might not work in general?

```{r}
tibble(json = got_chars) |>
  unnest_wider(json) |>
  select(id, where(is.list)) |>
  pivot_longer(
    where(is.list),
    names_to = "name",
    values_to = "value"
  ) |>
  unnest_longer(value)
```

5.  In `gmaps_cities`, what does `address_components` contain? Why does the length vary between rows? Unnest it appropriately to figure it out. (Hint: `types` always appears to contain two elements. Does [`unnest_wider()`](https://tidyr.tidyverse.org/reference/unnest_wider.html) make it easier to work with than [`unnest_longer()`](https://tidyr.tidyverse.org/reference/unnest_longer.html)?) .

# JSON

> JSON is short for **j**ava**s**cript **o**bject **n**otation and is the way that most web APIs return data. It’s important to understand it because while JSON and R’s data types are pretty similar, there isn’t a perfect 1-to-1 mapping, so it’s good to understand a bit about JSON if things go wrong.

## Data Types

It has six key data types. Four of them are scalars:

-   The simplest type is a null (`null`) which plays the same role as `NA` in R. It represents the absence of data.

-   A **string** is much like a string in R, but must always use double quotes.

-   A **number** is similar to R’s numbers: they can use integer (e.g., 123), decimal (e.g., 123.45), or scientific (e.g., 1.23e3) notation. JSON doesn’t support `Inf`, `-Inf`, or `NaN`.

-   A **boolean** is similar to R’s `TRUE` and `FALSE`, but uses lowercase `true` and `false`.

To represent multiple values you need to use one of the two remaining types: arrays and objects.

-   Both arrays and objects are similar to lists in R; the difference is whether or not they’re named.

    -   An **array** is like an unnamed list, and is written with `[]`. For example `[1, 2, 3]` is an array containing 3 numbers, and `[null, 1, "string", false]` is an array that contains a null, a number, a string, and a boolean.

    -   An **object** is like a named list, and is written with [`{}`](https://rdrr.io/r/base/Paren.html). The names (keys in JSON terminology) are strings, so must be surrounded by quotes. For example, `{"x": 1, "y": 2}` is an object that maps `x` to 1 and `y` to 2.

## jsonlite

-   This package convert JSON into R data structures

-   [`read_json()`](https://jeroen.r-universe.dev/jsonlite/reference/read_json.html) and [`parse_json()`](https://jeroen.r-universe.dev/jsonlite/reference/read_json.html). 

```{r}
# A path to a json file inside the package:
gh_users_json()
#> [1] "/home/runner/work/_temp/Library/repurrrsive/extdata/gh_users.json"

# Read it with read_json()
gh_users2 <- read_json(gh_users_json())

# Check it's the same as the data we were using previously
identical(gh_users, gh_users2)
#> [1] TRUE
```

```{r}
str(parse_json("1"))
```

```{r}
str(parse_json('[1, 2, 3]')) # array
```

```{r}
str(parse_json('{"x": [1, 2, 3]}')) # object
```

## Starting the rectangling process

> In most cases, JSON files contain a single top-level array

```{r}
json <- '[
  {"name": "John", "age": 34},
  {"name": "Susan", "age": 27}
]'

df <- tibble(json = parse_json(json))
df
```

```{r}
df |>
  unnest_wider(json)
```

> In rarer cases, the JSON file consists of a single top-level JSON object, representing one “thing”. In this case, you’ll need to kick off the rectangling process by wrapping it in a list, before you put it in a tibble.
```{r}
json <- '{
  "status": "OK", 
  "results": [
    {"name": "John", "age": 34},
    {"name": "Susan", "age": 27}
 ]
}
'
df <- tibble(json = list(parse_json(json)))
df
```

```{r}
df |>
  unnest_wider(json) |>
  unnest_longer(results) |>
  unnest_wider(results)
```

> Alternatively, you can reach inside the parsed JSON and start with the bit that you actually care about:

```{r}
df <- tibble(results = parse_json(json)$results)
df |>
  unnest_wider(results)
```

# Reference

[R for Data Science](https://r4ds.hadley.nz/rectangling.html)