---
title: "Rectangling Hierarchical Data"
subtitle: "R4DS Chapter 23 Concepts Applied to Google Analytics 4 in BigQuery"
author: "Jae Jung"
date: "March 8, 2026"
format:
  html:
    theme:
      light: flatly
      dark: darkly
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Contents"
    toc-expand: true
    code-fold: show
    code-copy: true
    code-tools: true
    highlight-style: github
    fig-width: 9
    fig-height: 5
    fig-align: center
    df-print: paged
    number-sections: true
    smooth-scroll: true
    embed-resources: true
    callout-appearance: minimal
execute:
  warning: false
  message: false
  cache: true
---

```{r}
#| label: setup
#| cache: false

library(bigrquery) # BigQuery connection & queries
library(DBI) # Standard database interface
library(dplyr) # Data manipulation
library(tidyr) # Rectangling: unnest_wider, unnest_longer, hoist
library(purrr) # Functional tools for list iteration
library(stringr) # String manipulation
library(lubridate) # Date/time handling
library(ggplot2) # Visualization
library(scales) # Axis formatting for ggplot2
library(gt) # Publication-quality tables
library(glue) # String interpolation
library(forcats) # Factor reordering for plots
library(tidyverse)

# Consistent ggplot2 theme for all charts
theme_set(
  theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 15),
      plot.subtitle = element_text(color = "grey40"),
      panel.grid.minor = element_blank()
    )
)
```

# Introduction {.unnumbered}

## What You Will Learn

This document bridges **R for Data Science Chapter 23 (Rectangling)** and the real-world complexity of **Google Analytics 4 (GA4) data** stored in BigQuery. You will learn to transform deeply nested, hierarchical event data into flat, analysis-ready tables — and then use those tables to answer genuine marketing questions.

::: callout-note
## Connection to R4DS Chapter 23

This document follows the structure of [R4DS Chapter 23: Hierarchical Data](https://r4ds.hadley.nz/rectangling.html). Every tool introduced there — `unnest_wider()`, `unnest_longer()`, and `hoist()` — appears here applied to real GA4 data. Where GA4 requires techniques **beyond** what Chapter 23 covers, those gaps are explicitly flagged and supplemented.
:::

## Learning Objectives

By the end of this workshop you will be able to:

1.  Explain why GA4 stores data in a hierarchical (nested) format in BigQuery
2.  Connect R to BigQuery using `{bigrquery}` and `{DBI}`
3.  Apply `unnest_wider()` to flatten GA4 **struct** columns (`device`, `geo`, `traffic_source`)
4.  Apply `unnest_longer()` to expand GA4 **repeated record** columns (`items`)
5.  Describe the **key-value struct pattern** used in `event_params` and explain why it requires a 3-stage rectangling pipeline not covered in R4DS Ch. 23
6.  Execute that 3-stage pipeline: `unnest_longer()` → `coalesce()` → `pivot_wider()`
7.  Produce marketing analyses for traffic sources, devices, content, and e-commerce using the Google Merchandise Store public dataset

## The Dataset: Google Merchandise Store (Obfuscated)

Google provides a **free public GA4 dataset** in BigQuery derived from their own online store ([merch.google.com](https://merch.google.com)). The data covers real shopping behaviour — page views, product interactions, and purchases — but user identifiers have been obfuscated for privacy.

| Detail | Value |
|--------------------|----------------------------------------------------|
| BigQuery path | `bigquery-public-data.ga4_obfuscated_sample_ecommerce` |
| Date range | November 2020 – January 2021 |
| Table format | One table per day: `events_YYYYMMDD` |
| Access | Free to read; your GCP project is billed for query processing |

------------------------------------------------------------------------

# The Challenge: Why Is GA4 Data Hierarchical?

## From Flat Files to Event Streams

Traditional web analytics tools exported flat CSV files: one row per session, one column per metric. GA4 takes a fundamentally different approach. It records every user interaction as an **event**, and because different event types carry completely different attributes, a flat structure is impractical.

Consider the variety:

-   A `page_view` event carries: page URL, page title, referrer, scroll depth
-   A `purchase` event carries: transaction ID, revenue, coupon codes, items purchased
-   A `click` event carries: link URL, link text, element ID

If these lived in a flat table, you would need hundreds of nullable columns. BigQuery's **nested and repeated field types** solve this elegantly — but require us to "rectangle" the data before we can analyse it.

## GA4's Three Nesting Patterns

GA4 data uses three distinct patterns of nesting. Understanding which pattern applies to each column is the key to choosing the right rectangling tool:

```{r}
#| label: nesting-patterns-table
#| echo: false

tibble::tribble(
  ~`Column(s)`                                      , ~`BigQuery Type` , ~`R List-Column Shape` , ~`Rectangling Tool` ,
  "device, geo,\ntraffic_source, app_info"          ,
  "RECORD (struct)"                                 ,
  "List of named lists\n(1 per row)"                ,
  "unnest_wider()"                                  ,
  "items"                                           ,
  "RECORD REPEATED\n(array of named structs)"       ,
  "List of tibbles with\nnamed columns"             ,
  "unnest_longer() +\nunnest_wider()"               ,
  "event_params,\nuser_properties"                  ,
  "RECORD REPEATED\n(array of key-value structs)"   ,
  "List of tibbles with\nkey + typed value columns" ,
  "unnest_longer() +\ncoalesce() +\npivot_wider()"
) |>
  gt() |>
  tab_header(title = "GA4 Nesting Patterns and Their Rectangling Tools") |>
  tab_style(
    style = cell_fill(color = "#f0f4ff"),
    locations = cells_body(rows = 3)
  ) |>
  tab_footnote(
    footnote = "Pattern 3 is not covered in R4DS Ch. 23 and is the main \
    supplementary topic of this document.",
    locations = cells_body(columns = `Rectangling Tool`, rows = 3)
  ) |>
  opt_stylize(style = 1)
```

::: callout-tip
## Visual Schema

Below is a simplified view of one row of GA4 data. Each indented block represents a nested structure:

```         
── event_date:         "20210101"          ← flat string
── event_name:         "purchase"          ← flat string
── user_pseudo_id:     "abc123..."         ← flat string
──┬ device:                                ← STRUCT (Pattern 1)
  ├── category:        "mobile"
  └── browser:         "Chrome"
──┬ geo:                                   ← STRUCT (Pattern 1)
  ├── country:         "United States"
  └── city:            "New York"
──┬ items: [...]                           ← ARRAY of STRUCTS (Pattern 2)
  ├── [0] item_id: "GGOEGAAX0104"
  │       item_name: "Google Tee"
  │       price: 22.99
  └── [1] item_id: "GGOEAAAB081014"
          item_name: "Google Cap"
          price: 15.99
──┬ event_params: [...]                    ← ARRAY of KEY-VALUE STRUCTS (Pattern 3)
  ├── [0] key: "transaction_id"
  │       value.string_value: "T-54321"
  ├── [1] key: "value"
  │       value.float_value: 38.98
  └── [2] key: "session_id"
          value.int_value: 1234567890
```
:::

------------------------------------------------------------------------

# Setup: Connecting to BigQuery

## Authentication and Connection

::: callout-important
## Before You Begin

You need a **Google Cloud Platform (GCP) account** with a project that has billing enabled. Querying the public dataset is free in terms of data costs (you won't be charged for the data itself), but GCP bills for query processing. Small queries on this dataset cost fractions of a cent.

1.  Go to [console.cloud.google.com](https://console.cloud.google.com)
2.  Create a project (or use an existing one)
3.  Enable the BigQuery API for that project
4.  Copy your **Project ID** — you'll need it below
:::

```{r}
#| label: connection
#| eval: false

# ── Replace with your GCP Project ID ─────────────────────────────────────────
# Either set the environment variable: usethis::edit_r_environ()
# and add: GCP_PROJECT_ID=your-actual-project-id
# Or assign directly (less portable):
billing_project <- "GCP_PROJECT_ID"

# ── Authenticate ──────────────────────────────────────────────────────────────
# First run: opens a browser for OAuth. Subsequent runs: uses cached token.
# In Positron IDE, if the browser redirect fails, add: use_oob = TRUE
bq_auth()
# bq_auth(use_oob = TRUE)  # Uncomment if you see an OAuth redirect error

# ── Connect to the public GA4 dataset ─────────────────────────────────────────
con <- dbConnect(
  bigrquery::bigquery(),
  project = "bigquery-public-data",
  dataset = "ga4_obfuscated_sample_ecommerce",
  billing = billing_project
)

# Verify connection: list available tables
dbListTables(con)
```

###  My Own BigQuery Account: Do not use this

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: step3-connect
#| eval: false

billing_project <- "ga4-bigquery-cppmsdm"

con <- dbConnect(
  bigquery(),
  project = "bigquery-public-data",
  dataset = "ga4_obfuscated_sample_ecommerce",
  billing = billing_project
)
```

## Downloading and Caching the Data

GA4 tables are large. For this workshop we download **one week of events** (\~150,000 rows) once, then save locally to avoid re-querying BigQuery on every render.

```{r}
#| label: download-data
#| eval: false

# ── Run this chunk ONCE. It saves data locally. ───────────────────────────────
dir.create("data", showWarnings = FALSE)

query_result <- bq_project_query(
  billing_project,
  "SELECT
     event_date,
     event_timestamp,
     event_name,
     event_params,
     user_pseudo_id,
     device,
     geo,
     traffic_source,
     ecommerce,
     items
   FROM `bigquery-public-data.ga4_obfuscated_sample_ecommerce.events_*`
   WHERE _TABLE_SUFFIX BETWEEN '20210101' AND '20210107'"
)

ga4_raw <- query_result |>
  bq_table_download(
    bigint = "numeric", # Convert INT64 to R numeric (sufficient for most analyses)
    page_size = 1000 # Download in batches of 10k rows
  )

saveRDS(ga4_raw, "data/ga4_raw.rds")
message("Saved ", nrow(ga4_raw), " rows to data/ga4_raw.rds")
```

> **Note**: RDS stands for R Data Serialization

```{r}
#| label: load-data
#| cache: false

# ── Load from local cache (use this for all subsequent analyses) ──────────────
ga4_raw <- readRDS("data/ga4_raw.rds")
```

------------------------------------------------------------------------

# Exploring the Raw Data Structure

Before rectangling anything, let us understand what we are dealing with.

## First Glimpse

```{r}
#| label: first-glimpse

# Overall structure
glimpse(ga4_raw)
```

> Notice that several columns show `<list>` as their type. These are **list-columns** — R's representation of BigQuery's nested fields. Each cell in a list-column contains a complex R object (a list, a tibble, or a named vector) rather than a single atomic value.

## Inspecting Individual List-Columns

Let us look inside each type of nested column:

```{r}
#| label: inspect-device
#| code-summary: "Inspect the device struct column"

# device is a STRUCT — each cell is a named list
ga4_raw$device[[1]] # Look at the first row's device record
```

```{r}
#| label: inspect-items
#| code-summary: "Inspect the items array column"

# items is an ARRAY of STRUCTs — each cell is a tibble (may be empty for non-purchase events)
# Find a row that has items (a purchase event)
purchase_rows <- which(ga4_raw$event_name == "purchase")
ga4_raw$items[[purchase_rows[1]]] # First purchase event's items
```

```{r}
#| label: inspect-event-params
#| code-summary: "Inspect the event_params key-value column"

# event_params is an ARRAY of KEY-VALUE STRUCTs
# Each cell is a tibble with columns: key, value (and value sub-columns)
page_view_rows <- which(ga4_raw$event_name == "page_view")
ga4_raw$event_params[[page_view_rows[1]]] # First page_view event's params
```

::: callout-note
## What You're Seeing

-   **`device`**: A single named list — one value per field. This is `Pattern 1`.
-   **`items`**: A tibble with one row per purchased item. Columns are the item's attributes (`item_name`, `price`, etc.). This is `Pattern 2`.
-   **`event_params`**: A tibble with one row per parameter, but instead of named attribute columns, it has just `key` and `value`. The `value` is itself a mini-struct with four typed sub-fields. This is `Pattern 3`— the most complex case.
:::

## Counting List-Column Lengths

```{r}
#| label: list-column-lengths

ga4_raw |>
  mutate(
    n_params = map_int(event_params, nrow),
    n_items = map_int(items, \(x) {
      if (is.null(x) || nrow(x) == 0) {
        0L
      } else {
        nrow(x)
      }
    })
  ) |> #select(n_params, n_items)
  group_by(event_name) |>
  summarise(
    n_events = n(),
    avg_params = mean(n_params),
    max_params = max(n_params),
    events_w_items = sum(n_items > 0),
    .groups = "drop"
  ) |>
  arrange(desc(n_events)) |>
  slice_head(n = 10) |>
  gt() |>
  tab_header(title = "Event Types: Complexity of Nested Fields") |>
  fmt_number(columns = avg_params, decimals = 1) |>
  cols_label(
    event_name = "Event Name",
    n_events = "# Events",
    avg_params = "Avg. Params",
    max_params = "Max Params",
    events_w_items = "Events with Items"
  ) |>
  opt_stylize(style = 1)
```

------------------------------------------------------------------------

# Part 1: Applying R4DS Rectangling to GA4 {#part1}

::: callout-note
## R4DS Chapter 23 Coverage

The techniques in this section are **directly taught in R4DS Chapter 23**. If you have worked through that chapter, this section applies those tools to real GA4 data. If not, this is your introduction to the rectangling toolkit.

**Key functions**: `unnest_wider()`, `unnest_longer()`, `hoist()` (all from `{tidyr}`)
:::

## What Is Rectangling?

**Rectangling** is the process of transforming a deeply nested hierarchical data structure (like a list-column) into a flat, rectangular data frame where every column contains atomic values and every row represents one observation.

The core insight from R4DS is that there are really only two fundamental operations:

| Operation | When to use | tidyr function |
|--------------------|------------------------|----------------------------|
| "Widen" a struct | Each list element is a **named list** (one record with named fields) | `unnest_wider()` |
| "Lengthen" an array | Each list element is a **vector/list of multiple things** (an array) | `unnest_longer()` |

Most complex cases combine both operations in sequence.

## Pattern 1: Simple Structs with `unnest_wider()` {#pattern1}

GA4's `device`, `geo`, and `traffic_source` columns are all **non-repeated RECORD types** — each row contains exactly one record with a fixed set of named fields. This is the simplest nesting pattern.

### The `device` Column

```{r}
#| label: unnest-device

device_data <- ga4_raw |>
  select(user_pseudo_id, event_name, event_date, device) |>
  unnest_wider(device) # ← expands named list into columns

glimpse(device_data) # 13 columns unnested
```

The `device` list-column is replaced by individual columns (13 of them): `category`, `mobile_brand_name`, `browser`, `operating_system`, etc. Each column name matches the original struct field name.

::: callout-tip
## `names_sep` for Avoiding Name Collisions

When unnesting multiple struct columns that might share field names, use `names_sep` to prefix each new column:

``` r
# Prevents collisions if two structs both have, e.g., a "version" field
ga4_raw |>
  unnest_wider(device, names_sep = "_") |>   # → device_category, device_browser
  unnest_wider(geo,    names_sep = "_")       # → geo_country, geo_city
```
:::

### The `geo` Column

```{r}
#| label: unnest-geo

geo_data <- ga4_raw |>
  select(user_pseudo_id, event_date, geo) |>
  unnest_wider(geo, names_sep = "_") # → geo_country, geo_city, geo_region

glimpse(geo_data)
```

### Unnesting Multiple Structs in One Pipeline

In practice, you usually want to unnest several struct columns at once:

```{r}
#| label: unnest-multiple-structs

ga4_flat_structs <- ga4_raw |>
  select(
    event_date,
    event_name,
    user_pseudo_id,
    event_timestamp,
    device,
    geo,
    traffic_source
  ) |>
  unnest_wider(device, names_sep = "_") |>
  unnest_wider(geo, names_sep = "_") |>
  unnest_wider(traffic_source, names_sep = "_")

# Now all struct columns are flat
names(ga4_flat_structs)
```

### Quick Visualisation: Device Category Mix

```{r}
#| label: device-category-chart

ga4_flat_structs |>
  count(device_category) |>
  mutate(
    pct = n / sum(n),
    label = glue("{device_category}\n{percent(pct, accuracy = 0.1)}")
  ) |>
  ggplot(aes(x = reorder(device_category, n), y = n, fill = device_category)) +
  geom_col(show.legend = FALSE, width = 0.6) +
  geom_text(aes(label = percent(pct, accuracy = 0.1)), hjust = -0.2, size = 4) +
  coord_flip() +
  scale_y_continuous(
    labels = comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  scale_fill_manual(values = c("#4285F4", "#34A853", "#FBBC05")) +
  labs(
    title = "Events by Device Category",
    subtitle = "Google Merchandise Store — Jan 1–7, 2021",
    x = NULL,
    y = "Number of Events"
  )
```

## Pattern 2: Arrays of Named Structs with `unnest_longer()` + `unnest_wider()` {#pattern2}

The `items` column contains **e-commerce line items**. Unlike `device`, this is a *repeated* RECORD — a purchase can have multiple items. Each row's `items` cell contains a tibble with one row per item.

This is the pattern covered in R4DS Chapter 23 using `gh_repos` as the example: an array where each element is a named struct.

### Ch23

-   When the list is unnamed, use `unnest_longer()`
-   When the list is name, use `unnest_wider()`

### Understanding the Two-Step Process

```{r}
#| label: items-structure

# Step 0: examine the items column for a purchase event
purchase_rows <- which(ga4_raw$event_name == "purchase")
ga4_raw |>
  filter(row_number() == purchase_rows[1]) |>
  select(event_name, items) |>
  unnest(items) # Quick look
```

-   Compare with below

```{r}
ga4_raw |>
  filter(row_number() == purchase_rows[1]) |>
  select(event_name, items) |>
  unnest_longer(items) |>
  unnest_wider(items)
```

> The pipeline for an array of named structs:

```         
items (list-column of tibbles)
  └─ unnest_longer(items)  → one row per item, items column now a single tibble-row
       └─ unnest_wider(items)  → each item field becomes its own column
```

```{r}
#| label: unnest-items

items_flat <- ga4_raw |>
  filter(event_name == "purchase") |>
  select(event_date, user_pseudo_id, event_timestamp, items) |>
  unnest_longer(items) |> # Step 1: one row per item
  unnest_wider(items) # Step 2: expand item struct fields into columns

glimpse(items_flat)
```

::: callout-note
## Why Two Steps?

`unnest_longer()` turns the *array* into individual rows (each element becomes its own row). After this step, the `items` column contains a single struct — which we then "widen" with `unnest_wider()`.

You can often shortcut with just `unnest(col)`, which does both steps for the common case of a list-column of data frames. But being explicit teaches the structure.
:::

### Top-Selling Products

```{r}
#| label: top-products

items_flat |>
  group_by(item_name) |>
  summarise(
    units_sold = sum(quantity, na.rm = TRUE),
    total_revenue = sum(item_revenue_in_usd, na.rm = TRUE),
    avg_price = mean(price_in_usd, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(!is.na(item_name)) |>
  arrange(desc(total_revenue)) |>
  slice_head(n = 10) |>
  gt() |>
  tab_header(
    title = "Top 10 Products by Revenue",
    subtitle = "Google Merchandise Store, Jan 1–7 2021"
  ) |>
  fmt_currency(columns = c(total_revenue, avg_price), currency = "USD") |>
  fmt_number(columns = units_sold, decimals = 0) |>
  cols_label(
    item_name = "Product",
    units_sold = "Units Sold",
    total_revenue = "Revenue",
    avg_price = "Avg. Price"
  ) |>
  opt_stylize(style = 1)
```

## Pattern 1b: `hoist()` for Selective Extraction {#hoist}

When you only need **a few fields** from a large struct, `hoist()` is more efficient than a full unnest because it avoids creating columns you don't need.

::: callout-note
## R4DS Coverage

`hoist()` is covered in R4DS Section 23.4. The syntax differs slightly from `unnest_wider()` — you name the output columns and specify the path to each value.
:::

```{r}
#| label: hoist-example

# Extract only country and device category without unnesting everything
ga4_targeted <- ga4_raw |>
  select(event_name, event_date, user_pseudo_id, device, geo) |>
  hoist(device, device_category = "category", browser = "browser") |>
  hoist(geo, country = "country", city = "city")

glimpse(ga4_targeted)
```

`hoist()` leaves the original list-columns in place (now with the extracted fields removed). This is useful when you want to keep the nested column for later processing while also having quick access to the most important fields.

------------------------------------------------------------------------

# Part 2: Beyond R4DS — The Key-Value Pattern {#part2}

::: callout-important
## GA4-Specific Content — Not in R4DS Ch. 23

This section covers a **pattern that R4DS Chapter 23 does not demonstrate**: the *key-value struct array*. This is the most complex and most important nesting pattern in GA4 data, used by `event_params` and `user_properties`.

Understanding this section is the central skill for GA4 analytics in R.
:::

## The Anatomy of `event_params`

In R4DS, the nested data examples (like `got_chars` or `gh_repos`) have **named fields** at each level. When you unnest a `gh_repo`, you get columns like `id`, `name`, `stargazers_count` — because those are the field names in the struct.

GA4's `event_params` is fundamentally different. Instead of named fields, it stores parameters in a **generic key-value format**:

```         
# What we WISH event_params looked like (named struct — covered by R4DS):
event_params = { page_location = "...", session_id = 123, engagement_time = 456 }

# What event_params ACTUALLY looks like (key-value struct — NOT in R4DS):
event_params = [
  { key = "page_location",   value = { string_value = "https://...", int_value = NULL, ... } },
  { key = "ga_session_id",   value = { string_value = NULL, int_value = 123,            ... } },
  { key = "engagement_time", value = { string_value = NULL, int_value = 456,            ... } }
]
```

**Why did Google design it this way?** Because different event types send completely different parameters. Using a generic `key-value` store means the schema never changes — new parameters can be added to any event without altering the table structure.

**The second complication**: the `value` field is itself a struct with four sub-fields — one for each data type that a parameter value might be:

```         
value.string_value  ← populated if the value is text
value.int_value     ← populated if the value is an integer
value.float_value   ← populated if the value is a decimal
value.double_value  ← populated if the value is a high-precision decimal
```

For any given parameter, exactly one of these will be non-NULL. We must *consolidate* them into a single value column.

## Step-by-Step: The Three-Stage Pipeline

### Toy Example — Building Intuition

Before applying to real data, let us build the intuition with a small hand-crafted example:

```{r}
#| label: toy-example
#| code-summary: "Toy example illustrating the key-value problem"

# Simulate what a single event's event_params looks like as a tibble
toy_params <- tibble(
  event_id = c(1L, 1L, 1L, 2L, 2L),
  event_name = c("page_view", "page_view", "page_view", "purchase", "purchase"),
  event_params = list(
    # event 1
    tibble(
      key = c("page_location", "ga_session_id", "engaged_session_event"),
      value = list(
        list(
          string_value = "https://shop.example.com/",
          int_value = NA,
          float_value = NA,
          double_value = NA
        ),
        list(
          string_value = NA,
          int_value = 7654321L,
          float_value = NA,
          double_value = NA
        ),
        list(
          string_value = NA,
          int_value = 1L,
          float_value = NA,
          double_value = NA
        )
      )
    ),
    # event 1 (second copy — same row)
    tibble(
      key = c("page_location", "ga_session_id", "engaged_session_event"),
      value = list(
        list(
          string_value = "https://shop.example.com/",
          int_value = NA,
          float_value = NA,
          double_value = NA
        ),
        list(
          string_value = NA,
          int_value = 7654321L,
          float_value = NA,
          double_value = NA
        ),
        list(
          string_value = NA,
          int_value = 1L,
          float_value = NA,
          double_value = NA
        )
      )
    ),
    # event 1 (third copy)
    tibble(
      key = c("page_location", "ga_session_id", "engaged_session_event"),
      value = list(
        list(
          string_value = "https://shop.example.com/",
          int_value = NA,
          float_value = NA,
          double_value = NA
        ),
        list(
          string_value = NA,
          int_value = 7654321L,
          float_value = NA,
          double_value = NA
        ),
        list(
          string_value = NA,
          int_value = 1L,
          float_value = NA,
          double_value = NA
        )
      )
    ),
    # event 2
    tibble(
      key = c("transaction_id", "value", "currency"),
      value = list(
        list(
          string_value = "T-00042",
          int_value = NA,
          float_value = NA,
          double_value = NA
        ),
        list(
          string_value = NA,
          int_value = NA,
          float_value = 49.98,
          double_value = NA
        ),
        list(
          string_value = "USD",
          int_value = NA,
          float_value = NA,
          double_value = NA
        )
      )
    ),
    # event 2 (second copy)
    tibble(
      key = c("transaction_id", "value", "currency"),
      value = list(
        list(
          string_value = "T-00042",
          int_value = NA,
          float_value = NA,
          double_value = NA
        ),
        list(
          string_value = NA,
          int_value = NA,
          float_value = 49.98,
          double_value = NA
        ),
        list(
          string_value = "USD",
          int_value = NA,
          float_value = NA,
          double_value = NA
        )
      )
    )
  )
)

cat("Starting shape:", nrow(toy_params), "rows x", ncol(toy_params), "cols\n")
cat("event_params column type:", class(toy_params$event_params), "\n")
toy_params |> select(event_id, event_name)
```

### Stage 1: `unnest_longer()` — Expand the Array

The first stage turns the array into rows. Each event's parameter list becomes individual rows, one per parameter:

```{r}
#| label: stage-1-unnest-longer

stage1 <- toy_params |>
  unnest_longer(event_params)

cat("After Stage 1:", nrow(stage1), "rows x", ncol(stage1), "cols\n")
stage1 |> select(event_id, event_name, event_params)
```

After this step, `event_params` is no longer a list of tibbles — it is now a single tibble-row per event parameter. But the `value` inside each row is still a nested named list.

### Stage 1b: `unnest_wider()` — Expand the Inner Struct

Now flatten the key and value struct fields into columns:

```{r}
#| label: stage-1b-unnest-wider

stage1b <- stage1 |>
  unnest_wider(event_params) |> # expands key, value into columns
  unnest_wider(value, names_sep = "_") |> # expands value struct sub-fields
  unnest_wider(value_1)
cat("After Stage 1b:", nrow(stage1b), "rows x", ncol(stage1b), "cols\n")
stage1b
```

We now have `key`, `string_value`, `int_value`, `float_value`, `double_value` as separate columns. But each row has only **one non-NULL value** across the four typed columns.

### Stage 2: `coalesce()` — Consolidate the Value Types

::: callout-important
## This Step Is Not in R4DS

`coalesce()` returns the first non-NULL value across its arguments. Here we use it to merge the four typed value columns into a single `value` column. This is a GA4-specific requirement not demonstrated in R4DS Chapter 23.
:::

```{r}
#| label: stage-2-coalesce

stage2 <- stage1b |>
  mutate(
    value = coalesce(
      string_value,
      as.character(int_value),
      as.character(float_value),
      as.character(double_value)
    )
  ) |>
  select(event_id, event_name, key, value) # Keep only what we need

cat("After Stage 2:", nrow(stage2), "rows x", ncol(stage2), "cols\n")
stage2
```

Now we have a clean long-format table: one row per parameter per event, with a single `value` column.

### Stage 3: `pivot_wider()` — Reshape to Analytical Format

The final step converts from long format (one row per parameter) to wide format (one column per parameter), giving us the flat, analysis-ready structure:

```{r}
#| label: stage-3-pivot-wider

stage3 <- stage2 |>
  pivot_wider(
    names_from = key,
    values_from = value,
    values_fn = first # If a key appears more than once, take the first
  )

cat("After Stage 3:", nrow(stage3), "rows x", ncol(stage3), "cols\n")
stage3
```

We now have one row per event and one column per parameter — exactly the flat structure we need for analysis.

## Applying the Full Pipeline to Real GA4 Data

Now let us apply the same three-stage pipeline to the real `ga4_raw` data:

```{r}
#| label: full-pipeline

event_params_wide <- ga4_raw |>
  # Keep only the columns we need before unnesting
  select(
    event_date,
    event_name,
    user_pseudo_id,
    event_timestamp,
    event_params
  ) |>
  unnest_longer(event_params) |> #stage 1
  unnest_wider(event_params) # stage 1b
  
event_params_wide_flat <- event_params_wide |> 
  unnest_wider(value, names_sep = "_") |>  # → value_string_value, value_int_value, etc.
  unnest_wider(value_1)

  # ── Stage 2: Consolidate the four typed value columns into one ─────────────
 event_params_wide_flat <- event_params_wide_flat |> 
  mutate(
    param_value = coalesce(
      string_value,
      as.character(int_value),
      as.character(float_value),
      as.character(double_value)
    )
  ) |>
  select(
    event_date,
    event_name,
    user_pseudo_id,
    event_timestamp,
    key,
    param_value
  ) |>

  # ── Stage 3: Pivot from long to wide (one column per parameter) ────────────
  pivot_wider(
    names_from = key,
    values_from = param_value,
    values_fn = first
  )

glimpse(event_params_wide_flat)
```

::: callout-tip
## What Columns Did We Get?

The resulting columns are all the `event_params` keys that appeared in the data. Common ones you'll see:

| Column                 | Description                                  |
|------------------------|----------------------------------------------|
| `page_location`        | Full URL of the viewed page                  |
| `page_referrer`        | URL that referred the user                   |
| `ga_session_id`        | Session identifier (integer)                 |
| `ga_session_number`    | Which session number this is for the user    |
| `session_engaged`      | `"1"` if the session met engagement criteria |
| `transaction_id`       | Purchase transaction ID                      |
| `value`                | Monetary value of the event                  |
| `currency`             | Currency code                                |
| `engagement_time_msec` | Milliseconds of engagement                   |
:::

## Discovering All Available Event Parameters

Before diving into analysis, it helps to know what parameters are available:

```{r}
#| label: param-inventory

param_inventory <- ga4_raw |>
  select(event_name, event_params) |>
  unnest_longer(event_params) |>
  unnest_wider(event_params) |>
  count(event_name, key, name = "n_occurrences", sort = TRUE) |>
  group_by(key) |>
  mutate(total = sum(n_occurrences)) |>
  ungroup()

# Most common parameters overall
param_inventory |>
  group_by(key) |>
  summarise(total_occurrences = sum(n_occurrences), .groups = "drop") |>
  arrange(desc(total_occurrences)) |>
  slice_head(n = 20) |>
  gt() |>
  tab_header(
    title = "Top 20 Most Common event_params Keys",
    subtitle = "Helps you know which parameters are available for analysis"
  ) |>
  fmt_number(columns = total_occurrences, decimals = 0) |>
  cols_label(
    key = "Parameter Key",
    total_occurrences = "Total Occurrences"
  ) |>
  opt_stylize(style = 1)
```

```{r}
#| label: param-by-event-type
#| code-summary: "Which parameters appear for which event types?"

# Heatmap-style table: parameters × event types
param_inventory |>
  filter(
    # Focus on the most analytically useful event types
    event_name %in%
      c(
        "page_view",
        "purchase",
        "add_to_cart",
        "view_item",
        "begin_checkout",
        "session_start"
      ),
    # And the most common parameters
    key %in%
      c(
        "page_location",
        "ga_session_id",
        "ga_session_number",
        "session_engaged",
        "transaction_id",
        "value",
        "currency",
        "engagement_time_msec",
        "page_referrer",
        "item_id",
        "item_name"
      )
  ) |>
  mutate(present = if_else(n_occurrences > 0, "✓", "")) |>
  select(key, event_name, present) |>
  pivot_wider(
    names_from = event_name,
    values_from = present,
    values_fill = ""
  ) |>
  gt() |>
  tab_header(
    title = "Parameter Availability by Event Type",
    subtitle = "✓ = parameter present for this event type"
  ) |>
  cols_label(key = "Parameter") |>
  opt_stylize(style = 1)
```

## `user_properties`: Same Pattern, Different Column

GA4's `user_properties` column follows **the same key-value struct pattern** as `event_params`. The only difference is what the keys represent — instead of event-level attributes, they capture persistent user-level attributes like `user_engagement_type`, `lifetime_value_bundle`, or custom dimensions you define in the GA4 interface.

```{r}
#| label: user-properties-pattern

# Confirm the structure is identical to event_params
ga4_raw |>
  filter(!map_lgl(user_properties, is.null)) |>
  slice(1) |>
  pull(user_properties) |>
  pluck(1)
```

```{r}
#| label: flatten-user-properties
#| code-summary: "Flatten user_properties using the same pipeline"

user_props_wide <- ga4_raw |>
  select(user_pseudo_id, event_date, user_properties) |>
  # Remove rows where user_properties is NULL or empty
  filter(!map_lgl(user_properties, \(x) is.null(x) || nrow(x) == 0)) |>
  distinct(user_pseudo_id, .keep_all = TRUE) |> # One record per user
  unnest_longer(user_properties) |>
  unnest_wider(user_properties) |>
  unnest_wider(value, names_sep = "_") |>
  mutate(
    prop_value = coalesce(
      value_string_value,
      as.character(value_int_value),
      as.character(value_float_value),
      as.character(value_double_value)
    )
  ) |>
  select(user_pseudo_id, event_date, key, prop_value) |>
  pivot_wider(
    names_from = key,
    values_from = prop_value,
    values_fn = first
  )

glimpse(user_props_wide)
```

::: callout-note
## Pattern 3 Summary

The **key-value struct pattern** requires three stages that build on each other:

```         
List-column of tibbles          (event_params before)
   │
   ▼  unnest_longer()           → Long format: one row per parameter
   │
   ▼  unnest_wider() ×2         → Columns for key + value sub-fields
   │
   ▼  coalesce()                → Single unified value column
   │
   ▼  pivot_wider()             → Wide format: one column per parameter
   │
Flat rectangular tibble         (event_params after)
```

This pipeline applies identically to `event_params` and `user_properties`.
:::

------------------------------------------------------------------------

# Reference

-   [R for Data Science, Ch23](https://r4ds.hadley.nz/rectangling.html)